A finite-state morphological transducer for Kyrgyz

### 2012-03-14 paper/paper.tex
ː

%Future TODO (was a FIXME): why copular suffixes?
%One result of making transducer targetted at machine translation is that we want to be able to both analyse, and generate the same forms. 
\subsection{Irregular negatives of finite verb forms}

@@

Since the general verbal negation morpheme in Kyrgyz tends to be /BA/ (as it is in all non-finite verb forms), we treated forms with /BA/ (the last two in the list above, for example) as regular (assuming other aspects of their morphology did not change).  We then created two different sets of continuation lexica for finite verb forms—one for regular finite verb forms, and one for irregular finite verb forms.  The continuation lexicon for regular finite verbal forms points to two continuation lexica: one of the regular verb suffixes (such as -/(I)птIр/ and -/E/, which in turn point to the appropriate continuation lexica for their person endings), and one of the negative -/BA/, which in turn points to the regular verb suffix continuation lexicon.  The continuation lexicon for irregular finite verb forms directly contains affirmative and negative morphology which each point to the appropriate personal suffix continuation lexica.

@@

%FIXME: maybe a footnote at least about what "optional loss of /D/" means?
%TODO: some explanation of how optional rules were dealt with
There are two rules that can immediately deal with some of these forms: optional loss of /D/ in the ablative suffix after 1st person singular, 2nd person singular, and 3rd person possession suffixes, and mandatory loss of /G/ in the dative suffix in the same situations.  Since these are rules specific to these morphological forms and do not apply generally in Kyrgyz at a phonological level, they were implemented directly in the interaction of the continuation lexica for these possessive suffixes and ablative and dative case suffixes.

@@

There were other vowel archiphonemes that had to be implemented in this project.  For example, \{U\} occurs in the past habitual suffix -/чU/ and the general gerund/infinitive -/Uː/, but nowhere else in the language.  Also, \{E\} was created for use in the habitual/future suffix; it has surface forms identical to those of \{A\}, except after vowels where it surfaces as [й].  Despite the fact that \{E\} is very similar to \{A\}, a separate but very similar twolc rule had to be created.  Repeated content could have been reduced by using cascading rules instead of two-level rules, but this would have caused other complications, such as finding the correct rule ordering.  An alternative possibility would have been to have a single intermediate level so that a rule like ``\{E\}→\{A\} after vowels'' with a default surface form of [й] could've been implemented.

@@

%FIXME: explain better
The orthography makes this straightforward, since e.g., words borrowed from Russian which end in ‹в› are treated in spoken Kyrgyz as unvoiced (and indeed have unvoiced surface forms syllable-finally), but are treated as voiced in the orthography.

@@

\subsection{Lenition}
In Kyrgyz, stem-final [voiceless] labial and dorsal consonants voice when a suffix beginning with a vowel follows them.  The implementation of this in twol is very straightforward.
%FIXME: what else needs to be said here?

@@

\begin{table}[htbp]
	\caption{Examples of /рн/-final nouns surfacing with no epenthetic vowel}\label{exrnnovowel}
\begin{center}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			%FIXME: should the glosses here and below be morpheme-by-morpheme for consistency, or is a more colloquial gloss okay?
			/мурн-(S)I/ & [мурду] & \eng{his/her/its nose} \\
			/мурн-Iм-DAн/ & [мурдуман] & \eng{from my nose} \\
			\bottomrule
		\end{tabular}
	}
	\hspace{0.5em}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/орн-(S)I/ & [орду] & \eng{his/her/its place} \\
			/орн-Iм-DAн/ & [ордуман] & \eng{from my place} \\
			\bottomrule
		\end{tabular}
	}
\end{center}
\end{table}

>>>>

%NONEEDFIX: why copular suffixes?
%One result of making transducer targetted at machine translation is that we want to be able to both analyse, and generate the same forms. 
\subsection{Irregular negatives of finite verb forms}

@@

Since the general verbal negation morpheme in Kyrgyz tends to be /BA/ (as it is in all non-finite verb forms), we treated forms with /BA/ (the last two in the table, for example) as regular (assuming other aspects of their morphology did not change).  We then created two different sets of continuation lexica for finite verb forms—one for regular finite verb forms, and one for irregular finite verb forms.  The continuation lexicon for regular finite verbal forms points to two continuation lexica: one of the regular verb suffixes (such as -/(I)птIр/ and -/E/, which in turn point to the appropriate continuation lexica for their person endings), and one of the negative -/BA/, which in turn points to the regular verb suffix continuation lexicon.  The continuation lexicon for irregular finite verb forms directly contains affirmative and negative morphology which each point to the appropriate personal suffix continuation lexica.

@@

There are two rules that can immediately deal with some of these forms: optional\footnote{Optional rules are dealt with by making a non-symmetrical transducer by way of marking lines in the lexc to be included in only one direction of the transducer; in this case, since the optional loss of /D/ is the default form, we do not generate forms where the /D/ remains, but do analyse them.} loss of /D/ in the ablative suffix after 1st person singular, 2nd person singular, and 3rd person possession suffixes, and mandatory loss of /G/ in the dative suffix in the same situations.  Since these are rules specific to these morphological forms and do not apply generally in Kyrgyz at a phonological level, they were implemented directly in the interaction of the continuation lexica for these possessive suffixes and ablative and dative case suffixes.

@@

There were other vowel archiphonemes that had to be implemented in this project.  For example, \{U\} occurs in the past habitual suffix -/чU/ and the general gerund/infinitive -/{\qipa Uː}/, but nowhere else in the language.  Also, \{E\} was created for use in the habitual/future suffix; it has surface forms identical to those of \{A\}, except after vowels where it surfaces as [й].  Despite the fact that \{E\} is very similar to \{A\}, a separate but very similar twolc rule had to be created.  Repeated content could have been reduced by using cascading rules instead of two-level rules, but this would have caused other complications, such as finding the correct rule ordering.  An alternative possibility would have been to have a single intermediate level so that a rule like ``\{E\}→\{A\} after vowels'' with a default surface form of [й] could've been implemented.

@@

%NONEEDFIX: explain better
The orthography makes idenifying voiced and voiceless pairs straightforward, since e.g., words borrowed from Russian which end in ‹в› are treated in spoken Kyrgyz as unvoiced (and indeed have unvoiced surface forms syllable-finally), but are treated as voiced in the orthography.

@@

\subsection{Lenition}
In Kyrgyz, stem-final [voiceless] labial and dorsal consonants voice when a suffix beginning with a vowel follows them.  The implementation of this in twol is very straightforward; our rules state that /п/ and /к/ become /б/ and /г/ preceding a morpheme boundary and when between two surface vowels with optional intervening non-surfacing phonemes.
%FIXED: does this sound better?

@@

\begin{table}[htbp]
	\caption{Examples of /рн/-final nouns surfacing with no epenthetic vowel}\label{exrnnovowel}
\begin{center}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			%FIXED: should the glosses here and below be morpheme-by-morpheme for consistency, or is a more colloquial gloss okay?
			/мурн-(S)I/ & [мурду] & \eng{his/her/its nose} \\
			/мурн-Iм-DAн/ & [мурдуман] & \eng{from my nose} \\
			\bottomrule
		\end{tabular}
	}
	\hspace{0.5em}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/орн-(S)I/ & [орду] & \eng{his/her/its place} \\
			/орн-Iм-DAн/ & [ордуман] & \eng{from my place} \\
			\bottomrule
		\end{tabular}
	}
\end{center}
\end{table}
